Seismic parameter estimation and the Canadian crust

### 2014-01-10 paper/bptectono.tex
recolor figures - changes in paper

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}

%% ------------------------------------------------------------------------ %%
\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. Existing models based on fractionating mantle composition or evolving mantle convection require accurate estimates of the geological and geophysical properties of crustal provinces to better understand the details of early continental formation (R. Durrheim, W. Mooney, 1991). Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of 1-D structure. Additionally, we investigate the feasibility of employing a variant on this method which allows resolution of $V_S$ and $V_P$ uniquely. Bootstrap error analysis is performed for each station dataset and subsets of these results are compared to the results for matching stations from previously published studies. Cross-analyzing these results with the mineral and rock seismic property database of Christensen (1996) will afford improved constraints on bulk geological composition of the Canadian landmass. These results will be used to make structural characterizations of the crust as well as evaluate competing models of crustal formation.
\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior form part of the assemblage that comprises the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly eastwest trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior are the extensive Interior Platforms comprising the southern extents of the Trans-Hudson Orogen and Hearne Domain as well as the Alberta Orogen. The Interior platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurasic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coastal mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and have enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Eaton (2005) who examines seismic evidence for several hypotheses of the origin of continental Moho: relic Moho, underplating, metamorphic front and regional décollement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis while the presence of high-velocity lower-crustal xenoliths in southern Alberta suggests thick mafic underplating and evidence for the second. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections seperated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional dècollement. Continuing from this effort, a study by Cook et al. (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust such as their observation that Achean cratons and their boundaries contain structures very similar to those found in Proterozoic and Phanerozoic orogens. Another key observation is the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors also remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary. The Lith5.0 revised crustal model presented by Perry, Eaton and Forte (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton.

@@

Despite diminishing support in recent years for the Conrad Discontinuity as a universal crustal feature, the division of the crust into seismically distinguishable upper and lower components remains a common assertion. In their review, Mooney and Brocher (1987) argued for a model of the continental crust separated intoa seismically reflective lower crust overlain by transparent upper crust. Their physical interpretation is that of a poorly laminated upper crust dominated by features of short length scale that contrasts with a lower crust composed of thin ($\sim$ 100 m) lamellae of significantly longer horizontal length scale. The inference of a laminated lower crust has since found support in studies of regions with extended or thicker, warmer crust (e.g. Ito, 1998; Mayer, 1996; Meissner et. al., 2006).

@@

The lower energy in the 20-30 km depth window noted in Section \ref{section:discontinuities} for the station profiles given in figure \ref{fig:discPCA} (left) provides some supporting evidence for seismically distinct upper and lower crustal layers. To better compare results across stations, we normalize each depth profile such that the Moho falls on a common ordinate (Figure \ref{fig:discInterp}, left) and stack the absolute values (Figure \ref{fig:discInterp}, right). The stacked signal has highest amplitude near the surface, where energy is associated with sedimentary layers, and at the Moho. There is a steady and marked decrease in energy over the bottom third of the profile that cannot be ascribed to sidelobe structure associated with the Moho pulse. This observation is ostensibly at odds with the Mooney and Brocher (1987) high reflectivity lower crustal model. Moreover, reflection profiles from {\sc Lithoprobe} transects (Hammer et. al., 2010) frequently display significant reflected energy through the crust with some transects including the Western Superior transect and the Cassiar in the Western Cordillera displaying pronounced reflectivity within the lower crust. We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength (Bostock, 1999). For 1 Hz teleseismic waves, crustal layering with scale lengths of 1000-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are limited to scale lengths under 100 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et .al. (1996) in their study of the Rhine Graben, have employed the term Conrad discontinuity to to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure 12, is a manifestation of such signals, but the lower energy levels ($<5\%$) in this mode would indicate that Conrad Discontinuity so-defined is at best weakly defined within the Canadian crust. This finding is supported by crustal models characterized by a first order continuously increasing velocity profile such as the gradual transition from a more silicic uppercrust to a more mafic lower crust in shields and platforms observed by Rolf Meisnner (Meisnner, 1986). It is also supported by the lack of evidence for a consistent mid-crustal reflection boundary in lithoprobe data (Hammer et. al., 2010). The increased transparency of the lower crust would appear to be better defined and would appear to relate to changes in rheology. Studies on crustal strength (Jackson, 2002) and on brittle and ductile models (Burov and Watts, 2006; Meisner and Rabbel, 2006) as well as the widely noted reduced seismicity of the lower crust indicate changes would support this contention, though see (Deichmann, 1991; Handy and Brun, 2004; Simpson, 1999) for competing evidence.

Mean crustal composition is another important parameter characterizing continental crust though it is important to distinguish it from actual lithological or mineralogical state of the crust. Rather, mean crustal composition is a representative average quantity that serves as a constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks (Rudnick and Gao, 2003). After the plate-tectonic revolution, Taylor (1967, 1977), proposed the 'island arc' or 'andesite' model which posited that bulk crustal chemical composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition (Christensen and Mooney, 1995; Rudnick and Fountain, 1995). Unlike these latter studies, which also utilize surface and exposed upper-crust geological information, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. (1981) though whereas Smithson utilizes mean crustal $V_P$ this study combines $V_P/V_S$ from teleseismic measurements with constraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.

@@

In Figure \ref{fig:lith} we plot our area-weighted seismic parameter estimates for the entire Canadian data set along with area-weighted average estimates for various tectonic subdivisions on a plot $V_P$ versus $V_S$. Also plotted are values for a selection of mineral assemblages considered to be representative of common crustal rocks from the compilation of laboratory measurements by Le Maitre (1976] and Christensen [1996), all quoted at average crustal pressures of 600 MPa. Dashed lines delimit isocontours in $V_P/V_S$. The choice of common shield lithologies is motivated by observations along 30km of paleo-depth represented in the Kapuskasing uplift of the Superior province (Percival and West, 1994). The Kapuskasing uplift comprises Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed from greenschist to amphibolite and granulite facies and early plutonic structures comprised largely of tonalites and their metamorphosed equivalents. Tonalites and granodiorites are also part of the TTG (tonalite-trondhjemite-granodiorite) suite which dominates Archaean cratons (Rudnick, 1995).

Seismic parameter estimates for Canada and the Canadian Shield fall, to a first-order approximation, between those of two candidate end-member lithologies: granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other earlier studies of bulk crustal composition e.g., (Rudnick and Fountain, 1995; Christensen and Mooney, 1995). The Slave, Superior and Churchill provinces possess comparable seismic parameters implying similar mean crustal compositions, with the Slave province represented as slightly more felsic in average composition. All three provinces cluster about the average value for the Canadian Shield. The Grenville Province, in contrast, displays markedly lower $V_S$ and higher $V_P/V_S$ implying a significant difference in mean composition. The high $V_P/V_S$ estimate, 1.77, is similar to the averaged results from a previous seismic study (Eaton et. al., 2005) that included stations of the central Grenville Orogen, and might be interpreted to indicate that the Grenville province possesses a mean crustal composition corresponding composition that is more mafic than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally (Eaton et. al., 2005; Rivers, 2008). However, many stations in the south-east of the province also display $V_P/V_S$ ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). Moreover, and more importantly, the average $V_P$ values for the Grenville province (a region well sampled by refraction profiles) requires low $V_S$ values that are inconsistent with a more mafic composition. Alternative explanations that allow for both high $V_P/V_S$ and reduced $V_P$, $V_S$ include the possibility that Grenville crust contains higher fluid concentrations than that of other tectonic provinces.

@@

Since our average seismic parameter estimates for the entire Canadian crust lie on a line connecting diorite with granite gneiss (\ref{fig:lith}), we may estimate bulk crustal composition in proportion to these lithologies as 51\% diorite and 49\% granite. A similar estimate for the Canadian Shield is 68\% diorite and 32\% granite. This lithological ratio is applied to the chemical compositions of diorite and granite using chemical averages calculated from 141 diorite and 197 granite samples compiled by Le Maitre (1976) to produce an estimate of bulk crustal composition. The resulting chemical breakdown is provided in table \ref{table:composition} along with previous estimates from global crustal studies. Our analysis reveal a crust with marginally higher average $SiO_2$ composition and lower $MgO$ content than these previous studies. Figure \ref{fig:composition} displays the major crustal element contributions from the data in table \ref{table:composition} normalized to our average Canadian crustal composition. The gray-shaded region indicates the $\pm 30\%$ range from our reference average Canadian crustal model.  Also included in this figure is the our average crustal composition estimate for the Canadian Shield. Its values closely track the global crustal estimates of Christensen \& Mooney (1995] and Smithson [1978). These data imply that the extensive platform and orogenic regions of Canada skew the average composition away from the global estimates of previous studies through incorporation of a higher proportion of granite gneiss. In contrast, the Canadian Shield crust aligns more closely with these previous global estimates.

Other researchers have employed Poisson's ratio (Zandt and Ammon, 1995) alone as an indicator of bulk crustal composition, and crustal thickness and {\it P}-wave velocity (Durrheim and Mooney, 1991) to constrain models of crustal formation. Zandt and Ammon (1995) report high $V_P/V_S$ values for Archean Shields and concluded that no systematic difference between Archean and Proterozoic crust exists. Their average, $V_P/V_S=1.84$, is much higher than our weighted average for the Canadian Shield of 1.74. These authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust, 1.73 compared to the corresponding value for our data of 1.77. They interpret these high values for older crust and low values for younger crust as favouring either of two limiting models. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This model implies that younger crust has yet to complete the full crustal genesis cycle. In their second model, Precambrian cratons evolve through unique processes and younger orogens involve the recycling and reworking of existing material with little net crustal growth.

Whereas Zandt and Ammon present evidence for increasing $V_P/V_S$ with crustal age, our estimates signify the opposite trend. Our results are more consistent with those of Durrheim and Mooney (1991) who argue that Proterozoic crust is thicker with higher {\it P}-velocity than Archean crust. The crustal thickness data presented in this study is only marginally thicker for Proterozoic crust (38.9 km) than Archean crust (38.0 km) and this difference is significantly smaller than the 5 km difference reported by Durrheim and Mooney. Though the difference is not large, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by Durrheim and Mooney (1991), has been persuasively augmented by recent work (Rudnick 1995; Rucknick and Gao, 2003) to now include a range of possible processes that may explain the data.

>>>>

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}
%% ------------------------------------------------------------------------ %%
\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurments are compared with results for matching stations from previous studies. Analysis of results in conjunction with additonal velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.
\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and has enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Eaton (2005) who considers seismic observations in light of several hypotheses for the origin of the continental Moho: relic Moho, underplating, metamorphic front and regional d\'ecollement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis whereas the presence of high-velocity, lower-crustal xenoliths in southern Alberta favours thick mafic underplating. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large-scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections separated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional d\'ecollement. A subsequent study by Cook et al. (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration.

In another recent summary study of Lithoprobe data, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust and note that Archean cratons and their boundaries contain structures similar to those found in Proterozoic and Phanerozoic orogens. They also discuss the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. (2009) and Hammer and Clowes (2011) raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary.

The Lith5.0 revised crustal model presented by Perry, Eaton and Forte (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick ($>$42.5km) Phanerozoic crust in the south and west of the continental craton.

@@

Despite diminishing support in recent years for the Conrad Discontinuity as a universal crustal feature, the division of the crust into seismically distinguishable upper and lower components remains a common assertion. In their review, Mooney and Brocher (1987) argued for a model of the continental crust separated into a seismically reflective lower crust overlain by transparent upper crust. Their physical interpretation is that of a poorly laminated upper crust dominated by features of short length scale that contrasts with a lower crust composed of thin ($\sim$ 100 m) lamellae of significantly longer horizontal length scale. The inference of a laminated lower crust has since found support in studies of regions with extended or thicker, warmer crust (e.g. Ito, 1998; Mayer, 1996; Meissner et. al., 2006).

@@

The lower energy in the 20-30 km depth window noted in Section \ref{section:discontinuities} for the station profiles given in figure \ref{fig:discPCA} (left) provides some supporting evidence for seismically distinct upper and lower crustal layers. To better compare results across stations, we normalize each depth profile such that the Moho falls on a common ordinate (Figure \ref{fig:discInterp}, left) and stack the absolute values (Figure \ref{fig:discInterp}, right). The stacked signal has highest amplitude near the surface, where energy is associated with sedimentary layers, and at the Moho. There is a steady and marked decrease in energy over the bottom third of the profile that cannot be ascribed to sidelobe structure associated with the Moho pulse. This observation is ostensibly at odds with the Mooney and Brocher (1987) high reflectivity lower crustal model. Moreover, reflection profiles from {\sc Lithoprobe} transects (Hammer et. al., 2010) frequently display significant reflected energy through the crust with some transects including the Western Superior transect and the Cassiar in the Western Cordillera displaying pronounced reflectivity within the lower crust. We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength (Bostock, 1999). For 1 Hz teleseismic waves, crustal layering with scale lengths of 1000-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are limited to scale lengths under 100 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et .al. (1996) in their study of the Rhine Graben, have employed the term Conrad discontinuity to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure 12, is a manifestation of such signals, but the lower energy levels ($<5\%$) in this mode would indicate that Conrad Discontinuity so-defined is at best weakly defined within the Canadian crust. This finding is supported by crustal models characterized by a first order continuously increasing velocity profile such as the gradual transition from a more silicic uppercrust to a more mafic lower crust in shields and platforms observed by Rolf Meisnner (Meisnner, 1986). It is also supported by the lack of evidence for a consistent mid-crustal reflection boundary in lithoprobe data (Hammer et. al., 2010). The increased transparency of the lower crust would appear to be better defined and relate to changes in rheology. Studies of crustal strength (Jackson, 2002) and deformation (Burov and Watts, 2006; Meisner and Rabbel, 2006) as well as the widely noted reduced seismicity of the lower crust indicate changes would support this contention; though see (Deichmann, 1991; Handy and Brun, 2004; Simpson, 1999) for competing evidence.

Mean crustal composition is another important parameter characterizing continental crust though it is important to distinguish it from actual lithological or mineralogical state of the crust. It serves as a representative, average quantity that affords constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks (Rudnick and Gao, 2003). After the plate-tectonic revolution, Taylor (1967, 1977), proposed the `island arc' or `andesite' model which posited that bulk crustal chemical composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition (Christensen and Mooney, 1995; Rudnick and Fountain, 1995). Unlike these latter studies, which also utilize surface and exposed upper-crust geological information, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. (1981) although, whereas Smithson utilizes mean crustal $V_P$, this study combines $V_P/V_S$ from teleseismic measurements with constraints on $V_P$ from previous active source studies as presented in the Crust 1.0 model.

@@

In Figure \ref{fig:lith} we plot our area-weighted seismic parameter estimates for the entire Canadian data set along with area-weighted average estimates for various tectonic subdivisions on a graph of $V_P$ versus $V_S$. Also plotted are values for a selection of mineral assemblages considered to be representative of common crustal rocks from the compilation of laboratory measurements by Le Maitre (1976) and Christensen (1996), all quoted at average crustal pressures of 600 MPa. Dashed lines delimit isocontours in $V_P/V_S$. The choice of common shield lithologies is motivated by observations along 30km of paleo-depth represented in the Kapuskasing uplift of the Superior province (Percival and West, 1994). The Kapuskasing uplift comprises Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed from greenschist to amphibolite and granulite facies and early plutonic structures comprised largely of tonalites and their metamorphosed equivalents. Tonalites and granodiorites are also part of the TTG (tonalite-trondhjemite-granodiorite) suite which dominates Archaean cratons (Rudnick, 1995).

Seismic parameter estimates for Canada and the Canadian Shield fall, to a first-order approximation, between those of two candidate end-member lithologies: granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other earlier studies of bulk crustal composition e.g., (Rudnick and Fountain, 1995; Christensen and Mooney, 1995). The Slave, Superior and Churchill provinces possess comparable seismic parameters implying similar mean crustal compositions, with the Slave province represented as slightly more felsic in average composition. All three provinces cluster about the average value for the Canadian Shield. The Grenville Province, in contrast, displays markedly lower $V_S$ and higher $V_P/V_S$ implying a significant difference in mean composition. The high $V_P/V_S$ estimate, 1.77, is similar to the averaged results from a previous seismic study (Eaton et. al., 2005) that included stations of the central Grenville Orogen, and might be interpreted to indicate that the Grenville province possesses a mean crustal composition corresponding composition that is more mafic than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally (Eaton et. al., 2005; Rivers, 2008). However, many stations in the south-east of the province also display $V_P/V_S$ ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). Moreover, and more importantly, the average $V_P$ value for the Grenville province (a region well sampled by refraction profiles) requires low $V_S$ values that are inconsistent with a more mafic composition. Alternative explanations that allow for both high $V_P/V_S$ and reduced $V_P$, $V_S$ include the possibility that Grenville crust contains higher fluid concentrations than that of other tectonic provinces.

@@

Since our average seismic parameter estimates for the entire Canadian crust lie on a line connecting diorite with granite gneiss (\ref{fig:lith}), we may estimate bulk crustal composition in proportion to these lithologies as 51\% diorite and 49\% granite. A similar estimate for the Canadian Shield is 68\% diorite and 32\% granite. This lithological ratio is applied to the chemical compositions of diorite and granite using chemical averages calculated from 141 diorite and 197 granite samples compiled by Le Maitre (1976) to produce an estimate of bulk crustal composition. The resulting chemical breakdown is provided in table \ref{table:composition} along with previous estimates from global crustal studies. Our analysis reveals a crust with marginally higher average $SiO_2$ composition and lower $MgO$ content than these previous studies. Figure \ref{fig:composition} displays the major crustal element contributions from the data in table \ref{table:composition} normalized to our average Canadian crustal composition. The gray-shaded region indicates the $\pm 30\%$ range from our reference average Canadian crustal model.  Also included in this figure is our average crustal composition estimate for the Canadian Shield. Its values closely track the global crustal estimates of Christensen \& Mooney (1995) and Smithson (1978). These data imply that the extensive platform and orogenic regions of Canada skew the average composition away from the global estimates of previous studies through incorporation of a higher proportion of granite gneiss. In contrast, the Canadian Shield crust aligns more closely with these previous global estimates.

Other researchers have employed Poisson's ratio (Zandt and Ammon, 1995) alone as an indicator of bulk crustal composition, and crustal thickness and {\it P}-wave velocity (Durrheim and Mooney, 1991) to constrain models of crustal formation. Zandt and Ammon (1995) report high $V_P/V_S$ values for Archean Shields and concluded that no systematic difference between Archean and Proterozoic crust exists. Their average, $V_P/V_S=1.84$, is much higher than our weighted average for the Canadian Shield of 1.74. These authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust (1.73) compared to the corresponding value for our data of 1.77. They interpret these high values for older crust and low values for younger crust as favouring either of two limiting models. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This model implies that younger crust has yet to complete the full crustal genesis cycle. In their second model, Precambrian cratons evolve through unique processes and younger orogens involve the recycling and reworking of existing material with little net crustal growth.

Whereas Zandt and Ammon present evidence for increasing $V_P/V_S$ with crustal age, our estimates signify the opposite trend. Our results are more consistent with those of Durrheim and Mooney (1991) who argue that Proterozoic crust is thicker with higher {\it P}-velocity than Archean crust. The crustal thickness data presented in this study reveals no significant difference between Proterozoic crust (38.9 km) and Archean crust (38.0 km) and this difference is significantly smaller than the 5 km difference reported by Durrheim and Mooney. Though the difference is not large, the higher seismic velocity ratio in the Proterozoic units (1.76) versus the Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by Durrheim and Mooney (1991), has been persuasively augmented by recent work (Rudnick 1995; Rucknick and Gao, 2003) to now include a range of possible processes that may explain the data.
